---
title: "G18_species_distribution - SDM"
author: "Nikola Gersak"
date: '2023-01-24'
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages(c('raster', 'rgdal', 'dismo', 'rJava', 'rcpp))
library(raster)
library(rgdal)
library(dismo)
library(tidyverse)
library(readxl)
library(sf)

```

# SPecies and habitats distribution predictions in Adriatic sea

## Data in general

Data that is being modelled concerns species in the Adriatic sea. Predictions are calculated for 4 types of species: *Coralium rubrum*, *Lithophaga Lithophaga*, *Scyllarides latus*, *Centrostephanus longispinus* and 2 types of habitats: sea reefs and caves.

First step is to load the data

Caution: for data wrangling and manipulation I'm using dplyr package and it's *%>%* pipeline function. 

```{r, data loading}

paths <- list.dirs(path = getwd(), full.names = TRUE, recursive = TRUE)
paths <- list.files(path = getwd())

path_to_data <- paste0(getwd(), '/', list.files(path = getwd())[1])
paths_data_files <- list.files(path = path_to_data, full.names = T)
paths_data_files_names <- list.files(path = path_to_data)

data <- map(setNames(paths_data_files, str_extract(paths_data_files_names, pattern = "\\d+")), function(x) read_excel(path = x, col_names = F))

grebeni <- read_excel(paths_data_files[7], sheet = 3)
splije <- read_excel(paths_data_files[7], sheet = 4)
SciLat <- read_excel(paths_data_files[7], sheet = 5)
CeLo <- read_excel(paths_data_files[7], sheet = 6)
LiLi <- read_excel(paths_data_files[7], sheet = 7)
CoRu <- read_excel(paths_data_files[7], sheet = 8)

```

a bit of wrangling

```{r}

grebeni <- grebeni %>% select(-c('Datum opažanja', Lokalitet, Napomena, 'Podatak unio')) %>% 
  mutate(bodovi = grebeni$`Vrednovanje strukture i funkcije morskih stanišnih tipova (bodovi)`,
         tip_vrst = grebeni$`Tipične vrste`,
         antr_utj = grebeni$`Antropogeni utjecaj`,
         x_cor = grebeni$`X koordinata`,
         y_cor =  grebeni$`Y koordinata`,
         inv_vrst = grebeni$`Invazivne vrste`,
         het_st = grebeni$`Heterogenost stanišnog tipa`,
         .keep = "used")


spilje <- spilje %>% select(-c('Datum opažanja', Lokalitet, Napomena, 'Podatak unio'))

```

## Raster files

Only two files of raster type were created and are containing data for predictions.

### Sea depth

Bathymetry data was obtained using the points batymethry data for Adriatic sea and interpolating it using thin plate spline regression in Saga gis (using QGIS interface).
Process can be repeated in r, using *Tps* package, and applying it's function to *bati_hr.shp* file.

### Sea habitats

Data was obtained by rasterizing the official Croatian habitats map of sea benthos.

Both rasters are made of pixels of 100m resolution.

## shp files

Vector data is more abundant and holds the information about the area where species and habitats occur as well as area where predictions of possible occurrence are made. Occurrence data is data obtained from the field inventories that were held throughout the year of 2022 and was put in the range tool to obtain the range data (see "EEA range tool"). The process of obtaining this data went in few steps:

1. vector features for 10x10 and 1x1 grid are taken from WFS and are used as base layer


```{r wfs loading, message=FALSE}

# URL of the WFS service
wfs_url <- 'http://services.bioportal.hr/wms'
wfs_request <- "request=GetFeature&service=WFS&version=2.0.0&typeName="
#load the sea benthos habitat map layer
wfs_layer <- "dzzpnpis:morski_bentos"
dsn <- paste0(wfs_url,"?", wfs_request, wfs_layer)
habitats_map <- st_read(dsn) %>% st_make_valid()
#load the 10x10 layer
wfs_layer <- "dzzpnpis:hr10x10"
dsn <- paste0(wfs_url,"?", wfs_request, wfs_layer)
layer_10x10 <- st_read(dsn) #%>% st_make_valid()
layer_10x10 <- layer_10x10[layer_10x10$geometry %overlaps% habitats_map$geometry, ]
#load the 1x1 layer
wfs_layer <- "dzzpnpis:hr1x1"
dsn <- paste0(wfs_url,"?", wfs_request, wfs_layer)
layer_1x1 <- st_read(dsn) %>% st_make_valid()
layer_1x1 <- layer_1x1[layer_1x1$geometry %overlaps% habitats_map$geometry, ]

```

2. zonal statistics for each square feature are calculated (mean and median)

Zonal statistics for rasters

```{r, rasterstats}
library(rasterStats)
#zonal statistics for raster file
mean_statistics <- zonal(r, grid_polygons, fun = mean)
median_statistics <- zonal(r, grid_polygons, fun = median)
#back to polygons
grid_polygons$mean <- mean_statistics
grid_polygons$median <- median_statistics
```

Zonal statistics for habitat maps and area percentage of each habitat in each grid cell calculated. This will be the number that will be used later as predictor.

```{r, habitats stats}
# Zonal statistics for habitats
habitats_map <- read_sf("./data/morski_bentos.shp")

#we can already load the model input data here
grid <- "1x1"
grid <- "10x10"
path_zs <- paste0("./data/slope_", grid, ".shp")
path_tv <- paste0("./data/training_slope_", grid, ".shp")

predictors <- read_sf(path_zs)

predictors_inersect <- st_intersection(predictors, habitats_map)
predictors_inersect$area <- st_area(predictors_inersect) %>% as.double
predictors_inersect$area_percentage <- predictors_inersect$area / as.double(st_area(predictors)) %>% as.double

# create data frame with the results
habitat_preferences <- data.frame(fid = predictors_inersect$fid,
                      habitat_id = predictors_inersect$nks_kod,
                      area = predictors_inersect$area,
                      area_percentage = predictors_inersect$area_percentage)

# pivot the data frame to have one column for each habitat feature
habitat_preferences <- habitat_preferences %>% pivot_wider(id_cols = c("fid"),
                                         names_from = "habitat_id", 
                                         values_from = "area_percentage")

habitat_preferences <- habitat_preferences %>% 
  rowwise() %>% 
  mutate_all(~round(sum(.),4))

```


3. zonal statistics are joined by location with all the calculated ranges from the range tool
4. no data values are added by hand in qgis to mark the locations that are definitely not places where the sea species and habitats can occur.
5. habitat preferences (weights) for species are added

```{r}
# Assign weights for each species and habitat differently
habitat_preferences <- habitat_preferences %>%
  mutate(area_pct_species_1001 = G.3.6. + G.4.3.,
         area_pct_species_1090 = G.3.6. + 0.5 * G.3.5. + G.4.3.,
         area_pct_species_1008 = G.3.6. + G.4.3. + G.3.2. + G.4.2.,
         area_pct_species_1027 = G.3.6.,
         area_pct_species_8330 = G.3.6. + G.4.3.,
         area_pct_species_1170 = G.3.6. + G.3.5. + G.4.3.)


#habitat preferences, weights of 0 and 1
habitat_preferences <- habitat_preferences %>% 
  mutate(habitat_preferences_1001 = unlist(list(ifelse(G.3.6. > 0 | G.4.3. > 0, 1, 0)))) %>%
  mutate(habitat_preferences_1090 = unlist(list(ifelse(G.3.6. > 0 | G.3.5. > 0, 1, 0)))) %>%
  mutate(habitat_preferences_1008 = unlist(list(ifelse(G.3.6. > 0 | G.4.3. > 0 | G.3.2. > 0 || G.4.2. > 0, 1, 0)))) %>%
  mutate(habitat_preferences_1027 = unlist(list(ifelse(G.3.6. > 0, 1, 0)))) %>%
  mutate(habitat_preferences_8330 = unlist(list(ifelse(G.3.6. > 0 | G.4.3. > 0, 1, 0)))) %>%
  mutate(habitat_preferences_1170 = unlist(list(ifelse(G.3.6. > 0 | G.3.5. > 0, 1, 0))))

#bathymethry preferences
habitat_preferences$bathymetry_preferred_1001 <- ifelse(predictors$bati_mean <= -30 & predictors$bati_mean >= -200, 1, 0)
habitat_preferences$bathymetry_preferred_1008 <- ifelse(predictors$bati_mean <= -40 & predictors$bati_mean >= -200, 1, 0)
habitat_preferences$bathymetry_preferred_1027 <- ifelse(predictors$bati_mean <= -1 & predictors$bati_mean >= -20, 1, 0)
habitat_preferences$bathymetry_preferred_1090 <- ifelse(predictors$bati_mean <= -4 & predictors$bati_mean >= -100, 1, 0)

```


last but not least, x and y coordinate variables are added, bacause y variable is later used as one of the predictors. Also, latitude preferences are added


```{r, data for model, warning=FALSE}
#loading the model input data
grid <- "1x1"
grid <- "10x10"
#path_zs <- paste0("./data/zonal_statistics_", grid, ".shp")
#path_tv <- paste0("./data/training_variables_", grid, ".shp")
path_zs <- paste0("./data/slope_", grid, ".shp")
path_tv <- paste0("./data/training_slope_", grid, ".shp")

predictors <- read_sf(path_zs) %>% 
  mutate(x = st_coordinates(st_centroid(.))[,1],
         y = st_coordinates(st_centroid(.))[,2]) %>% 
  as_data_frame()

#latitude preferences
habitat_preferences$lat_preferred_1008 <- ifelse(predictors$y <= 4900000, 1,
                                           ifelse(predictors$y <= 4931360, 0.5,
                                                  ifelse(predictors$y <= 4959120, 0.3,
                                                         ifelse(predictors$y > 4959120, 0.1, 0))))

habitat_preferences$lat_preferred_1090 <- ifelse(predictors$y <= 4900000, 1,
                                           ifelse(predictors$y <= 4931360, 0.5,
                                                  ifelse(predictors$y <= 4959120, 0.3,
                                                         ifelse(predictors$y > 4959120, 0.1, 0))))
#now we can join this in one table
predictors %>% left_join(habitat_preferences) -> predictors

train <- read_sf(path_tv) %>%   
  mutate(x = st_coordinates(st_centroid(.))[,1],
         y = st_coordinates(st_centroid(.))[,2]) %>% 
  as_data_frame()

train %>% left_join(habitat_preferences) -> train

train %>% select(Code) %>% unique %>% drop_na %>% unlist -> species_habitat_codes

#function for min max normalisation
min_max_norm <- function(x) {
    (x - min(x)) / (max(x) - min(x))
}
#normalisations
train %>% mutate(y_norm = min_max_norm(y),
                 bati_norm = min_max_norm(bati_mean),
                 slope_norm = min_max_norm(slope_mean)) -> train
predictors %>% mutate(y_norm = min_max_norm(y),
                      bati_norm = min_max_norm(bati_mean),
                      slope_norm = min_max_norm(slope_mean)) -> predictors

df_list <- list()

```

This shp_file can now go further in model

# Model and predictions

To keep the things simple, 4 predictor variables were used to predict one variable. General linear model is used to describe the relationship between these variables.


```{r, model, warning=FALSE}
#this for loop repeats the modelling process 6 times (=4 species, 2 habitats, needs to be repeated twice for 10x10 and 1x1 grid)
i <- "1090"
for (i in species_habitat_codes){
  species_habitat_code <- i

  train %>% select(fid, matches(species_habitat_code)) -> train_species
  train %>% select(-starts_with("species"), 
                   -starts_with("habitat"), 
                   -starts_with("bathymetry"),
                   -starts_with("G."),
                   -starts_with("area_pct"),
                   -starts_with("lat")) -> train_df
  #unique at the end is needed, otherwise it gives me "double column"
  train_df %>% left_join(train_species, by = "fid") %>% unique -> train_df
  
  train_df %>% select(-cellcode, -x, -y, 
                      -morska_mea, -morska_med, 
                      -layer, -path, -geometry, 
                      -bati_mean, -bati_media,
                      -slope_mean, -slope_medi) -> train_df
  #train_df %>% select(species_1090) %>% min_max_norm() -> train_df
  #na values in the Code variable are made up values
  train_df %>% 
    select(-fid) %>% 
    filter(Code == species_habitat_code | is.na(Code)) %>% 
    select(-Code) %>% 
    replace(is.na(.), 0) -> train_df
  
  #model
  #model <- glm(formula = pojava~., train_df, family = "binomial")
  if (species_habitat_code == "1090" | species_habitat_code=="1008"){
    formula <- paste0("pojava ~ bati_norm:bathymetry_preferred_", species_habitat_code, "+
                    y_norm:lat_preferred_", species_habitat_code, "+
                    area_pct_species_", species_habitat_code, ":habitat_preferences_", species_habitat_code)  
  }
  if (species_habitat_code == "1027" | species_habitat_code=="1001"){
    formula <- paste0("pojava ~ bati_norm:bathymetry_preferred_", species_habitat_code, "+
                    area_pct_species_", species_habitat_code)  
  }
  #habitats 8330 and 1170
  if (species_habitat_code=="1170"){
    formula <- paste0("pojava ~ bati_norm+slope_norm+
                    area_pct_species_", species_habitat_code)
   }
  if (species_habitat_code == "8330"){
  formula <- paste0("pojava ~ bati_norm+
                    area_pct_species_", species_habitat_code)
   }
  
  model <- glm(formula = as.formula(formula), data = train_df, family = "binomial")
  
  #the prediction part
  #also, I need to exclude the data that is already in predictors (subset = train_df)
  predictors$predicted <- predict(model, newdata = predictors)
  
  # Use a threshold to restrict predictions for 1090 and 1027 to central and south Dalmatia
  threshold <- 4928506
  if (species_habitat_code == "1090" | species_habitat_code=="1008"){
      predictors$predicted[predictors$y > threshold] <- 0
  }
  
  #saving results in the same variable
  predictors %>% mutate(pred_norm = min_max_norm(predicted),
                        round_p1 = round(predicted),
                        round_p2 = round(min_max_norm(predicted))) -> predictors
  
  #export
  #favourable reference range

  if (species_habitat_code == "1090" | species_habitat_code=="1008"){
    predictors %>% 
      filter(pred_norm > 0.2) %>% 
      select(-x,-y) %>% 
      st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_FRR.shp"), append = F)
    #favourable reference population or area
    FRV <- ifelse(species_habitat_code == "8330" | species_habitat_code == "1170", FRV <- "FRA", FRV <- "FRP")
    
    predictors %>% 
      filter(pred_norm > 0.4) %>% 
      select(-x,-y) %>% 
      st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_",FRV,".shp"), append = F)
  }
  if (species_habitat_code == "1001" | species_habitat_code=="1027"){
    predictors %>% 
    filter(pred_norm > 0.05) %>% 
    select(-x,-y) %>% 
    st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_FRR.shp"), append = F)
    #favourable reference population or area
    FRV <- ifelse(species_habitat_code == "8330" | species_habitat_code == "1170", FRV <- "FRA", FRV <- "FRP")
    predictors %>% 
      filter(pred_norm > 0.1) %>% 
      select(-x,-y) %>% 
      st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_",FRV,".shp"), append = F)
  }
    if (species_habitat_code=="1170"){
    predictors %>% 
    filter(pred_norm > 0.3) %>% 
    select(-x,-y) %>% 
    st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_FRR.shp"), append = F)
    #favourable reference population or area
    FRV <- ifelse(species_habitat_code == "8330" | species_habitat_code == "1170", FRV <- "FRA", FRV <- "FRP")
    predictors %>% 
      filter(pred_norm > 0.5) %>% 
      select(-x,-y) %>% 
      st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_",FRV,".shp"), append = F)
    }
    if (species_habitat_code == "8330"){
      predictors %>% 
        filter(pred_norm > 0.01) %>% 
        select(-x,-y) %>% 
        st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_FRR.shp"), append = F)
    #favourable reference population or area
      FRV <- ifelse(species_habitat_code == "8330" | species_habitat_code == "1170", FRV <- "FRA", FRV <- "FRP")
      predictors %>% 
        filter(pred_norm > 0.1) %>% 
        select(-x,-y) %>% 
        st_write(paste0("./exported_data/",species_habitat_code,"_",grid,"_predicted_",FRV,".shp"), append = F)
    }


  df_list[[i]] <- predictors
}

```

# Summary

Now that I've done the predictions, summary of statistics is left to do for each shapefile in folder where they were exported.

```{r}

folder <- "./exported_data"
shapefiles <- list.files(folder, pattern = "\\.shp$", full.names = TRUE)

sf_list <- lapply(shapefiles, function(x) {
  sf <- st_read(x)
  sf$file <- basename(x)
  return(sf)
})

sf_df <- bind_rows(sf_list)

sf_df <- sf_df %>%
  mutate(area = st_area(geometry)) %>%
  separate(file, c("id", "dim", "predicted", "type"), "_")

sf_df <- sf_df %>%
  mutate(species = case_when(
    id == "1001" ~ "Coralium rubrum",
    id == "1008" ~ "Centrostephanus longispinus",
    id == "1027" ~ "Lithophaga lithophaga",
    id == "1090" ~ "Scyllarides latus",
    id == "1170" ~ "Grebeni",
    id == "8330" ~ "Preplavljene ili dijelom preplavljene morske spilje",
    TRUE ~ NA_character_),
 type = sub("\\.shp", "", type))

areas_summarized <- sf_df %>%
  mutate(area_km2 = round(area / 10^6)) %>%
  group_by(id, dim, type, species) %>%
  summarize(total_area = sum(area_km2),
            n_feat = n())

#export
areas_summarized_without_geometry <- areas_summarized %>% as_data_frame %>%  select(-geometry)
write_csv(areas_summarized_without_geometry, "areas_summarized.csv")
#export as shp
areas_summarized %>% st_write("areas_sumarized.shp", append = FALSE)

```

